---
title: "Predicting College Football Games"
subtitle: "{{< meta params.season >}} Season Predictions"
author: "Phil Henrickson"
format: 
  html:
    echo: false
    message: false
    warning: false
    number-sections: false
params:
    season: 2024
    week: 4
---

```{r}
#| include: false
# packages
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(quarto)
library(gt)
library(gtExtras)
library(yardstick)
library(patchwork)

# src code
tar_source("R")

```

```{r}
# data from targets
team_info <- tar_read("cfbd_team_info_tbl")
team_estimates <- tar_read("season_team_estimates")
team_scores <- tar_read("team_scores")
draws <- tar_read("games_draws")
sims <- tar_read("games_sims")
```

```{r}
#| message: false
# request game info from collegefootballdata
game_info <-
  cfbfastR::cfbd_game_info(params$season) |>
  as_tibble() |>
  adjust_team_names() 

# betting lines
betting_lines <-
  cfbfastR::cfbd_betting_lines(year = params$season) |>
  as_tibble()

# prepare predictions
game_predictions <-
  sims |>
  join_team_divisions(games = game_info) |>
  prepare_fcs_teams() |>
  left_join(
    game_info |>
      select(game_id, start_date, completed)
  ) |>
  add_team_scores(
    teams_data = team_scores
  )

# add betting lines to game predictions
game_predictions_and_betting_lines = 
  game_predictions |>
  join_betting_lines(betting = betting_lines)


```


```{r}
add_gt_formatting = function(tbl, ...) {
  
  tbl |>
    gt::opt_row_striping(row_striping = F) |>
    gt::tab_options(table.font.size = 14,
                    ...)
}

prepare_spread_predictions = function(data) {
  
  data |>
    mutate(
      my_prediction = case_when(
        pred_margin > 0 ~ paste(home_team, pred_margin, sep = " by "),
        pred_margin < 0 ~ paste(away_team, -1 * pred_margin, sep = " by ")
      ),
      actual = case_when(
        home_margin > 0 ~ paste(home_team, home_margin, sep = " by "),
        home_margin < 0 ~ paste(away_team, -1 *home_margin, sep = " by ")
      ),
      spread_prediction = case_when(
        spread < 0 ~ paste(home_team, -1* spread, sep = " by "),
        spread > 0 ~ paste(away_team, spread, sep = " by ")
      )
    ) |>
    select(season, 
           week, 
           home_team, 
           away_team, 
           actual,
           home_margin, 
           my_prediction, 
           my_margin = pred_margin,
           provider, 
           margin = spread_margin,
           prediction = spread_prediction
    ) |>
    pivot_wider(names_from = c("provider"), 
                values_from = c("margin", "prediction"))
}

gt_pred_color = function(tbl, columns, target_columns) {
  
  tbl |>
    gt::data_color(
      columns = columns,
      domain = c(-80, 80),
      method = "numeric",
      palette= c("orange", "white", "dodgerblue3"),
      target_columns = target_columns,
      direction = "column",
      na_color = "white"
    )
  
}

spread_predictions_tbl = function(data) {
  
  data |>
    gt_tbl() |>
    gt::cols_align(
      #    columns = -c(home_team, away_team),
      align = "center"
    ) |>
    gt::cols_merge(
      columns = c("home_team", "away_team"),
      pattern = "{1} vs {2}",
    ) |>
    gt::cols_width(
      season ~ px(75),
      week ~ px(75)
    ) |>
    gt::cols_label(
      home_team = "Game",
      season = "Season",
      actual = "Result",
      week = "Week",
      my_prediction = "Prediction",
      prediction_DraftKings = "DraftKings",
      prediction_Bovada = "Bovada",
      `prediction_ESPN Bet` = "ESPN Bet"
    ) |>
    gt::cols_hide(
      columns = c("my_margin", "home_margin", "margin_DraftKings", "margin_Bovada", "margin_ESPN Bet")
    ) |>
    gt_pred_color(
      columns = "my_margin",
      target_columns = "my_prediction"
    ) |>
    gt_pred_color(
      columns = "margin_DraftKings",
      target_columns = "prediction_DraftKings"
    ) |>
    gt_pred_color(
      columns = "margin_Bovada",
      target_columns = "prediction_Bovada"
    ) |>
    gt_pred_color(
      columns = "margin_ESPN Bet",
      target_columns = "prediction_ESPN Bet"
    ) |>
    gt_pred_color(
      columns = "home_margin",
      target_columns = "actual"
    ) |>
    gt::opt_interactive(
      use_filters = T,
      use_compact_mode = T,
      page_size_default = 15
    )
}

plot_vs_betting_lines = function(data) {
  
  data |>
    ggplot(aes(x = -pred_margin, y = spread_open, label = paste(home_team, away_team, sep = " vs "))) +
    geom_point() +
    geom_text(vjust = -1, size = 1.5, check_overlap = T) +
    geom_abline() +
    theme_cfb() +
    xlab("Estimated Spread") +
    ylab("Vegas Spread") +
    ggpubr::stat_cor(aes(label = ..r.label..)) +
    facet_wrap(paste(season, paste("Week", week)) ~ .)+
    coord_cartesian(xlim = c(-50, 50),
                    ylim = c(-50, 50))
}

betting_lines_tbl = function(data) {
  
  data |>
    gt_tbl() |>
    gt::cols_align(
      align = "center"
    ) |>
    gt::cols_label(
      season = "Season",
      week = "Week",
      home_team = "Home",
      away_team = "Away",
    ) |>
    gt::data_color(
      method = "numeric",
      columns = c("Actual", "Prediction", "Bovada", "DraftKings", "ESPN Bet"),
      domain = c(-80, 80),
      palette = c("orange", "white", "dodgerblue3"),
      na_color = "white"
    ) |>
    gt::opt_interactive(
      use_filters = T,
      use_compact_mode = T,
      page_size_default = 25
    )
}


```

This page displays predictions for games in the {{< meta params.week >}} college football season season. These predictions come from a model built on historical college football play by play and game data in order to simulate upcoming games. All data is from collegefootballdata.com.

::: {.callout-note collapse="false"}

## About the Model

These predictions come from a game prediction model trained on historical college football data. My game prediction model uses team-level estimates of offensive, defensive, and special teams efficiency from play by play data. I simulate each game thousands of times in order to estimate the home team's win probability and expected margin of victory (or defeat).

Notably, I *do not use betting lines or the predictions of other models as features in my game prediction model*. This allows me to directly compare my performance against that of other models and betting lines.

:::

# Week {{< meta params.week >}} Predictions

The following table displays estimates and predictions for the upcoming week's college football games.

::: {.callout-note collapse="false"}

## About the Table

**Quality** indicates the quality of the teams involved in the game. This is the (harmonic) mean of the two team’s power ratings and is scaled to range from 0 to 100, with 100 indicating a game between highly rated opponents.

**Interest** indicates how interesting a game is expected to be based on the quality of the teams and how competitive the game is expected to be. I assign every game a competitive rating using a quadratic formula applied to Pr(HomeWin), which I use along with Quality in a weighted average to produce the Interest score. This ranges from 0 to 100, where 100 indicates a very competitive game between two highly rated opponents.

**Pr(HomeWin)** is the percentage of times that the home team won the game across every simulation.

**Prediction** is the average margin of victory for the winning team across all simulations.


:::

```{r}
#| column: body-outset-right
game_predictions |>
  prepare_game_predictions() |>
  filter(week == params$week) |>
  arrange(desc(game_interest)) |>
  game_predictions_tbl() |>
  add_gt_formatting()
```

## Betting Lines

How do these predictions compare to betting lines? I compare my predicted spread against the opening spread from ESPN Bet.

```{r}
#| fig-width: 7.5
game_predictions_and_betting_lines |>
  filter(provider == 'ESPN Bet') |>
  filter(week == params$week) |>
  plot_vs_betting_lines()+
  labs(y = 'ESPN Bet Spread')

```

The following table displays my prediction for each game compared to opening betting lines from various providers.

```{r}
#| column: body-outset-right
game_predictions_and_betting_lines |>
  filter(week == params$week) |>
  prepare_spread_predictions() |>
  spread_predictions_tbl() |>
  add_gt_formatting()
```

# Results as of Week {{< meta params.week >}}

How has the model performed to date?

The model predicts the probability that the home team will win and their expected margin of victory (or defeat). I assess the model’s performance by assessing the accuracy of its predictions (win/loss), the logloss of its probabilities, and the mean absolute error of the predicted margin.

<!-- The following tables display how the model has performed so far this season. -->

```{r}

assess_games = function(data, groups = c('season', 'season_type', 'week')) {
  
  class_metrics = yardstick::metric_set(mn_log_loss, accuracy)
  reg_metrics = yardstick::metric_set(mae)
  
  # add number of games
  tmp = 
    data |> 
    group_by(across(any_of(groups))) |>
    mutate(games = n_distinct(game_id)) |>
    group_by(across(any_of(groups)), games)
  
  class = 
    tmp |>
    class_metrics(
      truth = home_win,
      estimate = home_pred,
      home_prob
    )
  
  reg = 
    tmp |>
    reg_metrics(
      truth = home_margin,
      estimate = pred_margin
    )
  
  bind_rows(class,
            reg) |>
    select(any_of(groups), games, .metric, .estimate) |>
    mutate_if(is.numeric, round, 3)
}

pivot_estimates = function(data) {
  data |>
    pivot_wider(
      names_from = c(".metric"),
      values_from = c(".estimate")
    )
}

results_by_week = 
  game_predictions |>
  filter(completed == T) |>
  assess_games() |>
  pivot_estimates() |>
  gt_tbl() |>
  gt::cols_align(align = c("center")) |>
  gt::tab_header(title = "Model Results by Week")

results_overall =
  game_predictions |>
  filter(completed == T) |>
  assess_games(groups = c("season", "season_type")) |>
  pivot_estimates() |>
  gt_tbl() |>
  gt::cols_align(align = c("center")) |>
  gt::tab_header(title = "Season Model Results")


```

The following plot shows my model's predicted margin of victory vs the actual margin of victory for all games so far this season.

```{r}
plot_games_margins = function(data) {
  
  data |>
    ggplot(
      aes(
        x=pred_margin, 
        y=home_margin, 
        color = correct, 
        label = paste(home_team, away_team, sep = " vs "))
    )+
    geom_point()+
    geom_text(vjust = -1, size = 2, check_overlap = T)+
    scale_color_manual(values = c("red", "dodgerblue3"))+
    theme_cfb()+
    geom_vline(xintercept = 0, linetype = 'dotted')+
    geom_hline(yintercept = 0, linetype = 'dotted')+
    coord_cartesian(
      xlim = c(-75, 75),
      ylim = c(-75, 75)
    )+
    xlab("Predicted Home Margin")+
    ylab("Actual Home Margin")+
    guides(color = 'none')
  
}

results_plot  = 
  game_predictions |>
  filter(completed == T) |>
  mutate(
    correct = case_when(
      home_win == home_pred ~ 'yes',
      home_win != home_pred ~ 'no')
  ) |>
  plot_games_margins()

```

```{r}

results_overall

```


```{r}
#| fig-width: 7.5
results_plot

```

```{r}

results_by_week

```


### Previous Predictions

The following table displays the model's predictions for previous weeks along with the actual result.

```{r}
#| column: body-outset-right
#| 
game_predictions |>
  prepare_game_predictions() |>
  filter(start_date <= Sys.Date() & !is.na(actual)) |>
  group_by(week) |>
  arrange(desc(week), desc(game_interest)) |>
  ungroup() |>
  game_predictions_tbl() |>
  add_gt_formatting()

```

## Against the Spread

How has the model performed against the spread? I compare my model's predictions against betting lines from various providers.

```{r}

add_spread_cover = function(data) {
  
  data |>
    mutate(
      pred_cover = case_when(
        spread_margin >0 & pred_margin >= spread_margin ~ 'yes',
        spread_margin <0 & pred_margin < spread_margin ~ 'yes',
        spread_margin >0 & pred_margin <= spread_margin ~ 'no',
        spread_margin <0 & pred_margin >= spread_margin ~ 'no'
      ),
      actual_cover = case_when(
        spread_margin >0 & home_margin >= spread_margin ~ 'yes',
        spread_margin <0 & home_margin < spread_margin ~ 'yes',
        spread_margin >0 & home_margin <= spread_margin ~ 'no',
        spread_margin <0 & home_margin >=spread_margin ~ 'no'
      ) 
    ) |>
    mutate(
      across(c(pred_cover, actual_cover), ~ factor(.x, levels = c("yes", "no")))
    )
}

spread_predictions = 
  game_predictions_and_betting_lines |>
  filter(completed == T) |>
  filter(!is.na(spread)) |>
  select(game_id, season, week, home_team, away_team, home_margin, pred_margin, provider, spread, spread_open, spread_margin) |>
  add_spread_cover()

spread_predictions |>
  group_by(season, provider, week) |>
  mutate(games = n_distinct(game_id)) |>
  group_by(season, provider, week, games) |>
  yardstick::accuracy(
    actual_cover,
    pred_cover
  ) |>
  select(season,provider, week, .metric, .estimate) |>
  pivot_wider(
    names_from = c("provider"),
    values_from = c(".estimate")
  ) |>
  mutate_if(is.numeric, round, 3) |>
  gt_tbl() |>
  gt::cols_align(
    align = "center"
  )

```

The following table displays my model's predictions to date compared to betting

```{r}
#| column: body-outset-right
spread_predictions |>
  arrange(desc(week)) |>
  prepare_spread_predictions() |>
  spread_predictions_tbl() |>
  add_gt_formatting()

```

# Upcoming Games

The following table displays predictions for all upcoming games based on the results of the season so far.

```{r}
#| column: body-outset-right
game_predictions |>
  prepare_game_predictions() |> 
  filter(is.na(actual)) |>
  group_by(week) |>
  arrange(desc(week), desc(game_interest)) |>
  ungroup() |>
  game_predictions_tbl() |>
  add_gt_formatting()

```

